### 2022-11-15 description.tex
Add marginal, update results

While statistical alignment is typically performed using pairwise hidden Markov
models, finite-state transducers (FSTs) \green{present themselves as a better
alternative}.
% too brief?
FSTs \green{also} have the ability to rigorously model molecular sequence
evolution in addition to well established methods for combining them in
different ways \parencite{bradley2007transducers}.
A powerful and versatile algorithm is composition, which consists of sending the
output of one FST as the input of a second FST.
The FST model implemented in COATi is designed by composing smaller FSTs, each
representing a specific process.

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model that allows stop codons with gaps that
can occur at any position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represents a pairwise alignment.
% The shortest path through the FST corresponds with the best alignment.
However, composing large FSTs is expensive and can be prohibitive \green{(specify
I talk about time)}.
Despite the existence of efficient C++ FST libraries (e.g. openFST
\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
a few thousand nucleotides.
\green{To solve this issue, the search for an optimal path (alignment)} is
reformulated as a dynamic programming problem.

>>>>

% Summarize pair-HMMs. Brief, comment typically used but move onto FST right away
Statistical alignment is typically performed using pairwise hidden Markov models
(pair-HMMs).
Pair-HMMs are computational machines that have the ability to rigorously model
molecular sequence evolution and can calculate the probability that two
sequences are related, represented $P(X, Y)$ \parencite{yoon_2009_hmm}.
However, a limitation of pair-HMMs is the ability to only model the evolution
of two related sequences from an unknown ancestor.

Finite-state transducers (FSTs) have the same benefits as pair-HMMS with the
additional ability to generate a descendant sequence given an ancestral one.
Properly weighted, an FST can calculate the probability that a sequence $Y$
(descendant) evolved from sequence $X$ (ancestor), represented $P(Y | X)$.
Furthermore, the existence of well-established algorithms for combining FSTs in
different ways \parencite{bradley2007transducers} allows the design of complex
models by combining simpler FSTs.
A powerful and versatile algorithm for comparative sequence analysis is
composition, which consists of sending the output of one FST into the input of a
second FST.
The FST model implemented in COATi is designed by composing smaller FSTs, each
representing a specific process.

Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
The substitution models available are Muse and Gaut \parencite{muse_gaut_1994}
(MG94) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (ECM).
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model that allows stop codons with gaps that
can occur at any position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). (a) The
    substitution FST encodes a 64x64 codon substitution model with 64 arcs from
    M to S. (b) The indel FST allows for insertions (I to U) and deletions
    (D to V). Contiguous insertions and deletions are always arranged for
    insertions to precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represent a pairwise alignment.
% To find the most \green{probable/likely} alignment we must find the shortest
% path through.
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
All FST operations including model development, composition, search for the
shortest path, and other optimization algorithms are performed using the C++
openFST library \parencite{allauzen2007openfst}.

% However, composing large FSTs is an expensive operation and can be prohibitive.
% Despite the existence of efficient C++ FST libraries (e.g. openFST
% \cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
% than a few thousand nucleotides each.
% To solve this issue, the search for an optimal path (alignment) is
% reformulated as a dynamic programming problem.

COATi also features a marginal substitution model with substitution probability
matrix

\[P'_{ijp} = \mathlarger{\mathlarger{\sum}}_{cod} \thickspace \begin{cases}
P(i \thinspace | \thinspace cod) \enspace \text{if} \enspace cod_p = j\\
0 \enspace \text{otherwise}
\end{cases} \]

Where $P'_{ijp}$ represents the probability that codon $i$ from the ancestor
sequence changes to nucleotide $j$ of the descendant sequence at position $p$
$\in \{0, 1, 2\}$ of the reading frame.
$P$ is the substitution probability matrix MG94 or ECM.
This model emphasizes the position where the substitution in a codon occurs and
helps restrict the effects of low quality data in the descendant sequence.
In combination with the indel FST alignment with the marginal model can be
implemented using dynamic programming, which results in a significant speed up
with similar accuracy.

% Evolution FST + Q matrix
% Marginal model? - Probably not
% DP - probably not
% 1 figure: FST model? - results table?

### 2022-11-15 introduction.tex
Add marginal, update results

% sequence alignment is important and heavily used
Sequence alignment is a fundamental task in bioinformatics and a cornerstone
step in comparative and functional genomic studies
\parencite{sequence_alignment_rosenberg_2009}.
% often seen as an ad hoc problem
Modern sequence analysis began with the heuristic homology algorithms of
Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
progressed to arrive at current aligners such as Clustal$\Omega$
\parencite{clustal_omega_sievers_2011}, MAFFT \parencite{mafft_katoh_2002}, or
PRANK \parencite{prank_loytynoja_2014}.
However, the alignment of molecular sequences is, in practice, often seen as a
tool and the alignment inference as an ad hoc problem
\parencite{morrison_MSA_2018}.

% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
A common strategy is to perform alignment inference in the amino acid space
\parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts such as frameshifts and early stop codons.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% errors in genomic data that lead to erroneous downstream analyses
Errors in genomic datasets can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.

% to address this, we present COATi
Here we present COATi, short for COdon-Aware Alignment Transducer, a statistical
pairwise aligner that incorporates codon substitution models and is
robust to artifacts present in genomic data.

>>>>

% sequence alignment is important and heavily used
Sequence alignment is a cornerstone step in bioinformatics
\parencite{sequence_alignment_rosenberg_2009}.
% errors in genomic data that lead to erroneous downstream analyses
Uncorrected errors in sequence alignment can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
In addition, a common strategy is to perform alignment inference in the amino
acid space \parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not
support frameshifts or lack as statistical model.

% to address this, we present COATi
To address this problem, we present COATi, short for COdon-Aware Alignment
Transducer, a statistical pairwise aligner that incorporates codon substitution
models and is robust to artifacts present in genomic data.

### 2022-11-15 results.tex
Add marginal, update results

Using 4000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011}, MACSE v2.06
\parencite{ranwez_macse_2011}, MAFFT v7.407 \parencite{mafft_katoh_2002}, and
PRANK v.170427 \parencite{prank_loytynoja_2014} together with COATi.

After downloading, 1660 alignments contained gaps identified by at least one
aligner.
We randomly introduced gap patterns extracted from all five methods into the
2340 initially ungapped sequence pairs to generate the true alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics.
% by estimating $k_s$ and $k_a$ statistics.

@@

COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments,
and had a higher positive (97.3\%) selection accuracy.
MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
average alignment error an order of magnitude larger than COATi and the accuracy
retrieving positive selection of 81.5\% and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
obtained similar results with an average alignment error two orders of magnitude
larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
The accuracy retrieving negative selection was similar across all five
aligners.

>>>>

% by estimating $k_s$ and $k_a$ statistics.

@@

Both models in COATi were significantly (p < $2.2$x$10^{-16}$) more accurate at
inferring simulated alignments compared to other aligners according to the
one-tailed Wilcoxon signed ranked test, with an average alignment error
($d_{seq}$) value of $1$x$10^{-3}$.
In addition, COATi produced more perfect alignments ($d_{seq}=0$), less imperfect
alignments, and more accurately retrieved events of positive selection
(fst model 91.9\%, marginal model 90.8\%) (Supplementary Table 1).
Compared to COATi, MACSE (allows frameshifts) and MAFFT (using DNA) had an
average alignment error an order of magnitude larger than COATi and a lower
accuracy retrieving events of positive selection at a rate of 81.5\%
and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) had
an average alignment error two orders of magnitude larger than COATi and a
87.3\% and 69.1\% accuracy retrieving events of positive selection, respectively.
The accuracy retrieving events of negative selection was similar across all five
aligners (97.7\% $\pm$ 1.67\%).
